---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# nipnTK: National Information Platforms for Nutrition (NiPN) data quality toolkit
[![CRAN](https://img.shields.io/cran/v/nipnTK.svg)](https://cran.r-project.org/package=nipnTK)
[![CRAN](https://img.shields.io/cran/l/nipnTK.svg)](https://CRAN.R-project.org/package=nipnTK)
[![CRAN](http://cranlogs.r-pkg.org/badges/nipnTK)](https://cran.r-project.org/package=nipnTK)
[![Travis-CI Build Status](https://travis-ci.org/ernestguevarra/nipnTK.svg?branch=master)](https://travis-ci.org/ernestguevarra/nipnTK)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/ernestguevarra/nipnTK?branch=master&svg=true)](https://ci.appveyor.com/project/ernestguevarra/nipnTK)


**National Information Platforms for Nutrition (NIPN)** is an initiative of the 
European Commission to provide support to countries to strengthen their 
information systems for nutrition and to improve the analysis of data so as to 
better inform the strategic decisions they are faced with to prevent malnutrition 
and its consequences.



## Installation

You can install nipnTK from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("ernestguevarra/nipnTK")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
## basic example code
```
