---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# nipnTK: National Information Platforms for Nutrition (NiPN) data quality toolkit
[![CRAN](https://img.shields.io/cran/v/nipnTK.svg)](https://cran.r-project.org/package=nipnTK)
[![CRAN](https://img.shields.io/cran/l/nipnTK.svg)](https://CRAN.R-project.org/package=nipnTK)
[![CRAN](http://cranlogs.r-pkg.org/badges/nipnTK)](https://cran.r-project.org/package=nipnTK)
[![Travis-CI Build Status](https://travis-ci.org/validmeasures/nipnTK.svg?branch=master)](https://travis-ci.org/validmeasures/nipnTK)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/validmeasures/nipnTK?branch=master&svg=true)](https://ci.appveyor.com/project/validmeasures/nipnTK)
[![codecov](https://codecov.io/gh/validmeasures/nipnTK/branch/master/graph/badge.svg)](https://codecov.io/gh/validmeasures/nipnTK)

[National Information Platforms for Nutrition (NIPN)](http://www.nipn-nutrition-platforms.org)
is an initiative of the European Commission to provide support to countries to 
strengthen their information systems for nutrition and to improve the analysis of 
data so as to better inform the strategic decisions they are faced with to prevent 
malnutrition and its consequences.

As part of this mandate, NiPN has commissioned work on the development of a toolkit
to assess the quality of various nutrition-specific and nutrition-related data.
This is a companion R package to the toolkit of practical analytical methods that 
can be applied to variables in datasets to assess their quality.

The focus of the toolkit is on data required to assess anthropometric status such 
as measurements of weight, height or length, MUAC, sex and age. The focus is on 
anthropometric status but many of presented methods could be applied to other 
types of data. NiPN may commission additional toolkits to examine other variables 
or other types of variables.


## Requirements

* R version 3.4 or higher

Extensive use is made of the R language and environment for statistical computing. 
This is a free and powerful data analysis system. R provides a very extensive 
language for working with data. This companion package has been written using 
only a small subset of the R language. Many of the data quality activities 
described in the toolkit are supported by R functions included in this package
that have been written specifically for this purpose. These simplify the assessment 
of the quality of data related to anthropometry and anthropometric indices.


## Installation

You can install the development version of nipnTK from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("validmeasures/nipnTK")
```
